---
title: "Module 2: R Foundations with the Tidyverse"
output: 
    learnr::tutorial:
        progressive: false
        allow_skip: true # TODO: consider keeping or removing
        css: css/learnrIO_styles.css
        theme: "flatly"
runtime: shiny_prerendered
description: "lesson covering the basic R concepts and data structures, data importing/exporting, wrangling, tidying, and visualization using the Tidyverse."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(learnr)
library(tidyverse)
library(janitor)

tutorial_options(
  exercise.blanks = "___",
  exercise.checker = gradethis::grade_learnr()
)

source(here::here("inst/tutorials/module_2/www/module_datasets.R"))
source(here::here("inst/tutorials/module_2/www/helpers.R"))

employee_survey <- readr::read_csv(here::here("inst/extdata/employee_survey.csv"))
employee_turnover <- readr::read_csv(here::here("inst/extdata/employee_turnover.csv"))
employee_full_data <- readr::read_csv(here::here("inst/extdata/employee_full_data.csv"))
# gradethis_setup()

table_types <- tibble(
  Type = c("Numeric", "Integer", "Character", "Logical", "Factor"),
  Description = c("Decimal numbers or fractions", "Whole numbers", "Any text that is enclosed in quotations", "TRUE or FALSE values", "Categorical data with a fixed set of possible values"),
  Example = c(0.05, 42, "Hello World!", "TRUE", "factor(c('apple', 'orange', 'banana'))")
)
```




## Introduction
```{r, child="lessons/2.0_lesson-overview.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Basic R Concepts{data-progressive=TRUE}
```{r, child="lessons/2.1_basic-R-concepts.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Basic R Concepts: Exercises {data-progressive=TRUE}
```{r, child="lessons/2.1e_exercises.Rmd"}

```

<!-- ==============================NEXT TOPIC=============================== -->

## Introduction to the Tidyverse {data-progressive=TRUE}
```{r, child="lessons/2.2_intro-tidyverse.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Importing and exporting data {data-progressive=TRUE}
```{r, child="lessons/2.3_importing-and-exporting.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Tidy Data {data-progressive=TRUE}
```{r, child="lessons/2.4_tidy.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Transforming Data {data-progressive=TRUE}
```{r, child="lessons/2.5_transform.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Working with Strings and Text Data {data-progressive=TRUE}
```{r, child="lessons/2.6_text.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Data Visualization {data-progressive=TRUE}
```{r, child="lessons/2.7_viz.Rmd"}

```


