---
title: "Module 2: R Foundations with the Tidyverse"
output: 
    learnr::tutorial:
        progressive: true
        css: css/learnrIO_styles.css
        theme: "flatly"
runtime: shiny_prerendered
description: "add description"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(learnr)
library(dplyr)

source(here::here("inst/tutorials/module_2/www/module_datasets.R"))
# gradethis_setup()
```


# learnrIO

## Lesson Overview

<!-- ==============================NEXT TOPIC=============================== -->

## Basic R concepts and data structures {data-progressive=TRUE}
```{r, child="lessons/2.1_basic-R-concepts.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Intro to the Tidyverse {data-progressive=TRUE}
```{r, child="lessons/2.2_intro-tidyverse.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Importing and exporting data {data-progressive=TRUE}
```{r, child="lessons/2.3_importing-and-exporting.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Data Manipulation {data-progressive=TRUE}
```{r, child="lessons/2.4_wrangling.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Data Transformation {data-progressive=TRUE}
```{r, child="lessons/2.5_tidy.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

##  {data-progressive=TRUE}

### Sub-topic 1 
this is a sub-topic

### Sub-topic 2 
This is another sub-topic

<!-- ==============================NEXT TOPIC=============================== -->

##  {data-progressive=TRUE}

### Sub-topic 1 
this is a sub-topic

### Sub-topic 2 
This is another sub-topic

<!-- ==============================NEXT TOPIC=============================== -->

##  {data-progressive=TRUE}

### Sub-topic 1 
this is a sub-topic

### Sub-topic 2 
This is another sub-topic
