---
title: "Module 1: Introduction to RStudio & Project Workflow for Data Analytics"
output: 
    learnr::tutorial:
        progressive: false
runtime: shiny_prerendered
description: "learn the fundamentals of RStudio, including its interface, basic features, and project workflow, to efficiently manage and analyze data."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(learnr)
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(haven)


```


# learnrIO 

## Lesson Overview
```{r, child="lessons/1.0_overview.Rmd"}
``` 


<!-- ==============================NEXT TOPIC=============================== -->


## Overview of data analytics in the IO Psych business context {data-progressive=TRUE}

### Sub-topic 1 
this is a sub-topic

### Sub-topic 2 
This is another sub-topic


<!-- ==============================NEXT TOPIC=============================== -->

## Introduction to R, RStudio, and Quarto {data-progressive=TRUE}

### What is R & RStudio
```{r, child="lessons/1.1_r-and-rstudio.Rmd"}
``` 

### Navigating RStudio
```{r, child="lessons/1.2_nav-rstudio.Rmd"}
``` 


### Intro to Quarto and Interactive Reports
```{r, child="lessons/1.3_quarto.Rmd"}
``` 

<!-- ==============================NEXT TOPIC=============================== -->

## Version control: Git and Github {data-progressive=TRUE}

### Sub-topic 1 
this is a sub-topic

### Sub-topic 2 
This is another sub-topic

<!-- ==============================NEXT TOPIC=============================== -->

## Workflow and Data Management using RStudio Projects {data-progressive=TRUE}

### Sub-topic 1 
this is a sub-topic

### Sub-topic 2 
This is another sub-topic

<!-- ==============================NEXT TOPIC=============================== -->

