---
title: "Module 3: Exploratory Data Analysis with R"
output: 
    learnr::tutorial:
        progressive: true
        css: css/learnrIO_styles.css
        theme: "flatly"
runtime: shiny_prerendered
description: "you'll learn to formulate effective data-driven questions, perform exploratory data analysis (EDA) with R and the Tidyverse, craft compelling narratives with your data, communicate results to a business audience, and consolidate your skills with a hands-on practice report using Quarto."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(learnr)
library(tidyverse)

source(here::here("inst/tutorials/module_3/www/module_3_datasets.R"))

employee_survey <- read_csv(here::here("inst/extdata/mod_3_employee_survey.csv"))
employee_data <- read_csv(here::here("inst/extdata/mod_3_employee_missing_data.csv"))
satisfaction_survey <- read_csv(here::here("inst/extdata/mod_3_satisfaction_survey.csv"))
turnover_data <- read_csv(here::here("inst/extdata/mod_3_turnover_data.csv"))
employee_survey2 <- read_csv(here::here("inst/extdata/mod_3_employee_survey2.csv"))
practice_report <- read_csv(here::here("inst/extdata/practice_report_data.csv"))
```


## Introduction
```{r, child="lessons/3.0_lesson-overview.Rmd"}

```

<!-- ==============================NEXT TOPIC=============================== -->

## Introduction to Exploratory Data Analysis (EDA) {data-progressive=TRUE}
```{r, child="lessons/3.1_intro-EDA.Rmd"}

```

<!-- ==============================NEXT TOPIC=============================== -->

## Asking the Right Questions {data-progressive=TRUE}
```{r, child="lessons/3.2_asking-questions.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## EDA Techniques using R and the Tidyverse {data-progressive=TRUE}
```{r, child="lessons/3.3_eda-techniques.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Storytelling Your Data with R {data-progressive=TRUE}
```{r, child="lessons/3.4_storytelling.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Communicating Results in a Business Context {data-progressive=TRUE}
```{r, child="lessons/3.5_communicating.Rmd"}

```


<!-- ==============================NEXT TOPIC=============================== -->

## Practice EDA Report {data-progressive=TRUE}
```{r, child="lessons/3.6_practice-report.Rmd"}

```
