---
title: "Tidy Data"
---

### Introduction
As IO Psychologists, we often deal with a lot of data, be it employee performance metrics, survey responses, or recruitment data. One of the key challenges we face is ensuring that this data is well-organized and easy to analyze. Enter the concept of "Tidy Data" - a term coined by Hadley Wickham, the creator of many of the tools in the tidyverse package in R.

According to Wickham, "Tidy datasets are all alike, but every messy dataset is messy in its own way." The idea is to organize your data in a consistent way that makes it easier to work with and analyze. In essence, tidy data is about structuring datasets to facilitate analysis.  

### Tidy Data Principles

Tidy data follows three fundamental rules:  

- Each variable must have its own column.  
- Each observation must have its own row.  
- Each value must have its own cell.  


[Placeholder for an image illustrating tidy data]


The advantages of tidy data are twofold. Firstly, having a consistent data structure makes it easier to learn the tools that work with it. Secondly, placing variables in columns allows R's vectorized nature to shine, making data transformation feel natural. By maintaining tidy datasets, you can streamline your data analysis process, make it easier to handle missing values, and enhance the readability and organization of your data.  

Let's consider an example where we have data on employee satisfaction, performance, and tenure. The tidy version of this data would look something like this:  

| Employee ID | Satisfaction | Performance | Tenure |
|---|---|---|---|
| 1 | 7 | 8 | 2 |
| 2 | 8 | 7 | 3 |
| 3 | 9 | 9 | 5 |
| 4 | 6 | 5 | 1 | 

In this example, each variable (Satisfaction, Performance, Tenure) has its own column, each observation (each employee) has its own row, and each value has its own cell.  


### Pivoting Data

In practice, you'll often encounter data that is not tidy. For instance, a variable might be spread across multiple columns, or one observation might be scattered across multiple rows. To handle such scenarios, `tidyr`, a core tidyverse package for working with messy data, provides two pivotal functions: `pivot_longer()` and `pivot_wider()`.  

Let's discuss each one in detail.  

### `pivot_longer()` 

is used to make "wide" data longer. In other words, it takes multiple columns and combines them into key-value pairs, duplicating all other columns as needed.  

```{r, eval=FALSE}
tidy_survey_data <- survey_data |> pivot_longer(
  cols = starts_with("Question"),
  names_to = "Question",
  values_to = "Answer"
)
```

`pivot_wider()` does the opposite. It makes "long" data wider, spreading out a key-value pair across multiple columns.  

```{r, eval=FALSE}
wide_survey_data <- tidy_survey_data |> pivot_wider(
  names_from = "Question",
  values_from = "Answer"
)
```


### Separate and unite
At times, we might have a single column containing two variables. In such a case, we can use the `separate()` function to split that column into two. For example, imagine a column 'Question_Answer' in our survey data that combines the question number and the answer.  

```{r, eval=FALSE}
tidy_survey_data <- untidy_survey_data |> separate(
  Question_Answer, 
  into = c("Question", "Answer"), 
  sep = "_"
)
```

Conversely, we might need to combine two columns into one. The unite() function does this.  
```{r, eval=FALSE}
untidy_survey_data <- too_tidy_survey_data |> unite(
  "Question_Answer", 
  Question, Answer, 
  sep = "_"
)
```


### Your Turn

Time to apply what you've learned about tidying data! Here are three exercises for you to try using the survey_data, untidy_survey_data, and too_tidy_survey_data datasets.  

**Exercise 1:** Use `pivot_longer()` on the `survey_data` dataset to transform it into a long format where we have three distinct columns: 'Employee', 'Question', and 'Answer'.  

```{r}

```


**Exercise 2:** The `untidy_survey_data` dataset has a 'Question_Answer' column that combines the question number and the answer. Use the `separate()` function to split this column into two separate columns: 'Question' and 'Answer'.  

```{r}

```


**Exercise 3:** The `too_tidy_survey_data` dataset has separated 'Question' and 'Answer' columns. Use the `unite()` function to combine these into a single column named 'Question_Answer'.  

```{r}

```




Remember to inspect the results after each step to ensure that your transformations have worked as expected. **Keep up the good work!**
