---
title: "Lesson Overview"
---
### Welcome to **Module 1: Introduction to RStudio & Quarto!**  

This lesson will serve as an introduction to R, RStudio, and Quarto. We will cover the basics of navigating and using R, RStudio, and Quarto for data analytics in the context of IO Psychology.  

This module's lessons are:  

- **Introduction to R and RStudio:** This lesson will provide an overview of RStudio, R, and Quarto, including how to download and install them. You will also learn the basics of R and RStudio, including how to navigate the RStudio interface, run code, and find help when you need it.  

- **Project Workflow and Reports using Quarto:** The final lesson will introduce you to the workflow of creating projects in RStudio and generating reports using Quarto. It will cover markdown basics, adding images and links, creating code chunks, understanding YAML, and knitting documents in different output formats.  


Before getting started, we **strongly** encourage you to create a Quarto file to work in alongside this course. An Quarto file is a great way to keep track of notes and practice what you learn from the course. By working in a Quarto file, you can write and execute R code, include text, images, and other media, and export your work to a variety of formats (such as HTML or PDF). This allows you to create a comprehensive record of your progress throughout the course.  

In this module, we will be using the following packages:  

```{r, eval=FALSE}
install.packages("tidyverse")
```


If you have not yet installed these packages, you can do so using the `install.packages()` function in R, as shown above. Once installed, you can load the packages into your current R session using the `library()` function:  

```{r, eval=FALSE}
library(tidyverse)
```


So, be sure to load these in your Quarto file!


Remember, learning R and data analytics is a journey. Don't worry if you don't grasp everything immediately. Practice and persistence are key. **Let's get started!**  
