---
title: "What is R and Rstudio?"
---

### Welcome to **Introduction to R and RStudio!**  

### Introduction to the R language

R is a powerful programming language that was designed with statistics and data analysis in mind. It has a rich ecosystem of packages which can be utilized to perform complex data manipulations, analyses and visualizations. Moreover, it is highly extensible and allows you to integrate with other languages (like C++, Python, Julia, etc.) if needed.  


### Introduction to RStudio
  
RStudio is an integrated development environment (IDE) for R, providing a convenient and user-friendly interface for writing, running, and debugging R code. It includes several useful features such as syntax highlighting, code completion, and version control integration. RStudio makes working with R easier and more enjoyable. RStudio is often used by I/O psychology practitioners because it simplifies many of the technical aspects of coding, allowing practitioners to focus more on the insights gleaned from the data.  

It is quite possible you may be thinking, "so, what's the difference between R and RStudio? Aren't they the same thing?" Well, imagine R as a versatile and powerful kitchen appliance like a food processor, while RStudio is like a cookbook filled with recipes and step-by-step instructions on how to use the appliance.


[insert image]


R is the underlying programming language and software environment for statistical computing and graphics. It's like a food processor that can help you perform various tasks such as slicing, chopping, and blending ingredients. R is what allows you to perform data analysis, create visualizations, and develop statistical models.

On the other hand, RStudio is an integrated development environment (IDE) designed to work with R. It's like a cookbook that teaches you how to use the food processor (R) effectively, guiding you through the process of preparing a variety of dishes. RStudio provides features such as code editing, debugging, and project management, making it easier to write, run, and organize your R code.


[insert image]


### Navigating RStudio

#### Pane Layout

RStudio's interface is divided into four main panes:

- **Script Editor:** This is where you will write and edit your R scripts. It features syntax highlighting and code completion.  

- **Console:** This is where R code is executed. You can also type code directly into the console, but it's often better to write code in the script editor and then run it, so you have a record of your code.  

- **Environment/History:** The Environment tab shows your current workspace, including all variables, data frames, and functions that exist in memory. The History tab shows your command history.  

- **Files/Plots/Packages/Help:** This pane allows you to navigate your file system, see plots generated by your code, manage packages, and access R's built-in help system.  

[Image Placeholder]  

<br>
#### Addins
<br>


#### Running Code in the Console

To run code from your script editor, you can simply highlight the code you want to run and press the `Run` button, or use the shortcut `Ctrl+Enter` (`Cmd+Enter` on Mac). The code will be sent to and executed in the console.  

Let's run a simple line of code:  

```{r}
print("Hello, I/O Psychology!")
```


After running this code, the console will display the message "Hello, I/O Psychology!".  


#### Environment Pane

The Environment pane in RStudio is an extremely helpful tool as it displays all the objects (like vectors, data frames, and functions) that you've created in your R session. You can see the object's name, class, length, and content, giving you a quick overview of what's currently in your workspace.  

Let's create a simple vector and a data frame to see how they appear in the Environment pane:  

```{r}
# create a vector of employee ages
employee_ages <- c(25, 30, 35, 40)

# create a data frame of employees
employees <- data.frame(
  id = 1:4,
  age = employee_ages,
  department = c("HR", "Marketing", "Sales", "IT")
)
```


[Image Placeholder]  


### Getting Help

As you start your journey with R, you will undoubtedly encounter challenges or unfamiliar tasks. That's perfectly normal! One of the most important skills in programming, and especially in data analysis, is knowing how to find help when you need it. Let's explore some of the ways you can get help.  


#### Built-in R Help System

R has a built-in help system that you can access directly from RStudio. If you want to know more about a function, you can use the `help()` function or the `?` shortcut, followed by the name of the function. For example, if we wanted to learn more about the `print()` function, we could use:  

```{r}
help(print)
# or 
?print
```


This will open a help page in the Files/Plots/Packages/Help pane, providing detailed information about the function.  


[Image Placeholder]  


#### Online Resources

While the built-in help system is great for learning about specific functions, you will often encounter more complex problems or need broader advice. That's where online resources come in handy. Here are a few key ones:  

- **Google:** Don't underestimate the power of a well-phrased Google search! Often, someone else has already encountered the same problem and found a solution. Try to be as specific as possible in your search, and include "R" to get more relevant results.  

- **Stack Overflow:** This is a question-and-answer website for a wide range of programming topics, including R. You can search for your problem, and if it hasn't been asked before, you can post a new question. Remember to be specific and include a minimal reproducible example if possible.  

- **GitHub:** Many R packages are developed openly on GitHub, and you can often find help and examples in the repository's issues or wiki. If you think you've found a bug in a package, you can report it in the issues section.  

- **Package Documentation and Vignettes:** Most R packages come with detailed documentation and vignettes (long-form guides). You can access the documentation by using the `help()` function or `?`, and vignettes can be accessed with the `vignette()` function.  

- **R-Bloggers:** This is an aggregate of blogs about R, where you can find tutorials, examples, and advice on a wide range of topics. [R-bloggers website](https://www.r-bloggers.com/).  

- **Twitter:** Many R users and developers are active on Twitter, using the hashtag #rstats. It's a great place to ask questions, share your work, and keep up with the latest developments in the R community.  

- **Reddit:** The Reddit community can also be a helpful resource, particularly the r/rstats and r/RStudio subreddits. You can post questions, share your work, and engage in discussions with other R users. As with other online communities, try to be specific in your questions and provide a reproducible example if possible. Remember, the goal is to make it as easy as possible for others to understand your problem and provide help.  


### Conclusion

Learning to effectively search for help and solutions online is a valuable skill that will greatly aid your journey into R and data analysis. The more you practice, the better you'll become at finding the information you need.  




Remember, everyone starts from scratch and it's okay not to know everything. The R community is known for being welcoming and helpful, so don't hesitate to ask questions when you're stuck!  

